---
title: "Supply Demand Gap Analysis"
output: pdf_document
date: "2024-09-28"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = dirname(rstudioapi::getSourceEditorContext()$path)) #cool little line to 
# set the working directory to wherever the rmd is saved, everyone should be able to run the rmd then. 

knitr::opts_chunk$set(echo = TRUE)

getwd()

#Library Load: install the following packages if you do not have them allready
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(readxl)
library(httr)
library(zoo)
```


```{r LMI Job Data Loading}
# This code reads the LMI jobs demand data by region and occupation from the ODJFS website. 

#Paths
common_path <- getwd()
target_folder <- paste0(common_path, "/data/lmi-data/")

# Create the target folder, this will be helpful so all groupmemebers will automatically have a folder created
  dir.create(target_folder, recursive = TRUE)

# URLs for the different regions, these are all the excell sheets on the Ohio LMI website for each region
url_northeast <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/Northeast.xlsx"
url_central <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/Central.xlsx"
url_west <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/West.xlsx"
url_southeast <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/Southeast.xlsx"
url_northwest <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/Northwest.xlsx"
url_southwest <- "https://ohiolmi.com/_docs/PROJ/JobsOhio/Southwest.xlsx"



# Process Northeast region First
temp_northeast <- tempfile(fileext = ".xlsx")
response_northeast <- GET(url_northeast, write_disk(temp_northeast, overwrite = TRUE)) #Calls the url, trys to read the xlsx
  headers_northeast <- suppressMessages(read_excel(temp_northeast, range = cell_rows(3:6))) 
  #Have to get rid of bad headers
  headers_northeast <- apply(headers_northeast, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_northeast <- c(headers_northeast, "med wage symbol")
  data_northeast <- suppressMessages(read_excel(temp_northeast, skip = 5)) 
  #Skip the first 5 rows! all headers of white space. 
  colnames(data_northeast) <- headers_northeast[1:12]
  #grab the names from only these headers
  rows_all_na_northeast <- rowSums(is.na(data_northeast)) == ncol(data_northeast)
  first_all_na_row_northeast <- which(rows_all_na_northeast)[1]
  data_northeast <- data_northeast[1:(first_all_na_row_northeast - 1), ]
  #that's annoying, but this should give us JUST the headers and not weird splits or missing headers. 
  data_northeast$jobsohioregion <- "Northeast"


#OKAY, now do the same thing for all the other 5 regions, just past the above and change the region name as appropriate. We could probably write a loop....but I am lazy
# Process Central region________________________________________________________
temp_central <- tempfile(fileext = ".xlsx")
response_central <- GET(url_central, write_disk(temp_central, overwrite = TRUE))
  headers_central <- suppressMessages(read_excel(temp_central, range = cell_rows(3:6)))
  headers_central <- apply(headers_central, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_central <- c(headers_central, "med wage symbol")
  data_central <- suppressMessages(read_excel(temp_central, skip = 5))
  colnames(data_central) <- headers_central[1:12]
  rows_all_na_central <- rowSums(is.na(data_central)) == ncol(data_central)
  first_all_na_row_central <- which(rows_all_na_central)[1]
    data_central <- data_central[1:(first_all_na_row_central - 1), ]
  data_central$jobsohioregion <- "Central"



# Process West region_______________________________________________________________
temp_west <- tempfile(fileext = ".xlsx")
response_west <- GET(url_west, write_disk(temp_west, overwrite = TRUE))
  headers_west <- suppressMessages(read_excel(temp_west, range = cell_rows(3:6)))
  headers_west <- apply(headers_west, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_west <- c(headers_west, "med wage symbol")
  data_west <- suppressMessages(read_excel(temp_west, skip = 5))
  colnames(data_west) <- headers_west[1:12]
  rows_all_na_west <- rowSums(is.na(data_west)) == ncol(data_west)
  first_all_na_row_west <- which(rows_all_na_west)[1]
    data_west <- data_west[1:(first_all_na_row_west - 1), ]
  data_west$jobsohioregion <- "West"



# Process Southeast region___________________________________________________________________
temp_southeast <- tempfile(fileext = ".xlsx")
response_southeast <- GET(url_southeast, write_disk(temp_southeast, overwrite = TRUE))
  headers_southeast <- suppressMessages(read_excel(temp_southeast, range = cell_rows(3:6)))
  headers_southeast <- apply(headers_southeast, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_southeast <- c(headers_southeast, "med wage symbol")
  data_southeast <- suppressMessages(read_excel(temp_southeast, skip = 5))
  colnames(data_southeast) <- headers_southeast[1:12]
  rows_all_na_southeast <- rowSums(is.na(data_southeast)) == ncol(data_southeast)
  first_all_na_row_southeast <- which(rows_all_na_southeast)[1]
    data_southeast <- data_southeast[1:(first_all_na_row_southeast - 1), ]
  data_southeast$jobsohioregion <- "Southeast"


# Process Northwest region___________________________________________________________________________
temp_northwest <- tempfile(fileext = ".xlsx")
response_northwest <- GET(url_northwest, write_disk(temp_northwest, overwrite = TRUE))
  headers_northwest <- suppressMessages(read_excel(temp_northwest, range = cell_rows(3:6)))
  headers_northwest <- apply(headers_northwest, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_northwest <- c(headers_northwest, "med wage symbol")
  data_northwest <- suppressMessages(read_excel(temp_northwest, skip = 5))
  colnames(data_northwest) <- headers_northwest[1:12]
  rows_all_na_northwest <- rowSums(is.na(data_northwest)) == ncol(data_northwest)
  first_all_na_row_northwest <- which(rows_all_na_northwest)[1]
    data_northwest <- data_northwest[1:(first_all_na_row_northwest - 1), ]
  data_northwest$jobsohioregion <- "Northwest"



# Process Southwest region____________________________________________________________________
temp_southwest <- tempfile(fileext = ".xlsx")
response_southwest <- GET(url_southwest, write_disk(temp_southwest, overwrite = TRUE))
  headers_southwest <- suppressMessages(read_excel(temp_southwest, range = cell_rows(3:6)))
  headers_southwest <- apply(headers_southwest, 2, function(x) paste(na.omit(x), collapse = " "))
  headers_southwest <- c(headers_southwest, "med wage symbol")
  data_southwest <- suppressMessages(read_excel(temp_southwest, skip = 5))
  colnames(data_southwest) <- headers_southwest[1:12]
  rows_all_na_southwest <- rowSums(is.na(data_southwest)) == ncol(data_southwest)
  first_all_na_row_southwest <- which(rows_all_na_southwest)[1]
    data_southwest <- data_southwest[1:(first_all_na_row_southwest - 1), ]
  data_southwest$jobsohioregion <- "Southwest"

# Combine all region datasets into a single data frame
lmi_oews <- bind_rows(data_northeast, data_central, data_west, data_southeast, data_northwest, data_southwest)
#OKAY! all Regions loaded.




#Ohio overall data_____________________________________________________________________________
# Define the column names manually, including the new 'median_wage_symbol'. This is because I cannot get the same method as the regions to work for Ohio overall, ran out of time. 
column_names <- c(
  "soc_code",                    # SOC Code
  "soc_lmi_title",               # Occupational Title
  "employment",                  # Employment* 2020 Annual
  "projected_2030",              # 2030 Projected
  "change_employment",           # Change in Employment 2020-2030
  "percent_change",              # Percent
  "annual_openings_growth",      # Annual Openings Growth
  "exits",                       # Exits
  "transfers",                   # Transfers
  "total_openings",              # Total
  "median_wage",                 # Median Wage May 2021
  "median_wage_symbol",          # med wage symbol
  "Typical Education Needed for Entry",    # Not used in the select list
  "Work Experience in a Related Occupation",    # Not used in the select list
  "Typical On-The-Job Training Needed to Attain Competency" # Not used in the select list
)


# Read the data from the Excel file, skipping the first three rows. I could not get the url to read in the same way....so I just downloaded this one. https://view.officeapps.live.com/op/view.aspx?src=https%3A%2F%2Fohiolmi.com%2F_docs%2FPROJ%2FOhio%2FOccOH30.xlsx&wdOrigin=BROWSELINK
ohio_data <- read_excel(paste0("./data/lmi-data/OccOH30_raw.xlsx"), 
                        sheet = "Occupational Detail", skip = 3, col_names = FALSE)
ohio_data <- as.data.frame(ohio_data)
# Assign the manually defined column names to the data, these are defined above
colnames(ohio_data) <- column_names
# Add a new column 'jobsohioregion' with all values set to 'Ohio', this will give us the same manually added data column as the previous region code-chunks
ohio_data <- ohio_data %>%
  mutate(jobsohioregion = 'Ohio')




#Combine Ohio and Region Data___________________________________________________________________________
# Ensure consistent column names and types for `ohio_data_trimmed`
ohio_data_trimmed <- ohio_data %>%
  select(
    soc_code, soc_lmi_title, employment, projected_2030, 
    change_employment, percent_change, annual_openings_growth, 
    exits, transfers, total_openings, median_wage, 
    median_wage_symbol, jobsohioregion
  ) %>%
  mutate(
    employment = as.numeric(employment),  # Convert to numeric
    change_employment = as.numeric(change_employment),  
    median_wage = as.numeric(median_wage), 
    projected_2030 = as.numeric(projected_2030), 
    percent_change = as.numeric(percent_change),
    annual_openings_growth = as.numeric(annual_openings_growth),
    exits = as.numeric(exits),
    transfers = as.numeric(transfers),
    total_openings = as.numeric(total_openings)
  )

# Ensure column names and types match for `lmi_oews`
lmi_oews <- lmi_oews %>%
  rename(
    soc_code = `SOC Code`,
    soc_lmi_title = `Occupational Title`,
    employment = `Employment* 2020 Annual`,
    projected_2030 = `2030 Projected`,
    change_employment = `Change in Employment 2020-2030`,
    percent_change = `Percent`,
    annual_openings_growth = `Annual Openings Growth`,
    exits = `Exits`,
    transfers = `Transfers`,
    total_openings = `Total`,
    median_wage = `Median Wage May 2021`,
    median_wage_symbol = `med wage symbol`
  ) %>% mutate(
    employment = as.numeric(employment),  # Convert to numeric
    projected_2030 = as.numeric(projected_2030),
    change_employment = as.numeric(change_employment),
    percent_change = as.numeric(percent_change),
    annual_openings_growth = as.numeric(annual_openings_growth),
    exits = as.numeric(exits),
    transfers = as.numeric(transfers),
    total_openings = as.numeric(total_openings),
    median_wage = as.numeric(median_wage)
  )

# Ensure standardized column names for both data frames
colnames(ohio_data_trimmed) <- tolower(trimws(colnames(ohio_data_trimmed)))
colnames(lmi_oews) <- tolower(trimws(colnames(lmi_oews)))

# Combine the two datasets
ohio_region_lmi_data <- bind_rows(lmi_oews, ohio_data_trimmed)%>%
  mutate(
    jobsohioregion = case_when( #casewhen easiest in this case
      jobsohioregion == "Northwest" ~ 1L,
       jobsohioregion == "West" ~ 2L,
      jobsohioregion == "Southwest" ~ 3L,
      jobsohioregion == "Northeast" ~ 4L,
      jobsohioregion == "Central" ~ 5L,
      jobsohioregion == "Southeast" ~ 6L,
      jobsohioregion == "Ohio" ~ 39L, #ohio to 39, check this is true for all
      TRUE ~ NA_integer_  # For any unmatched regions, set to NA, should removed these or see why they occured
     )
    )
#Will have to fix manual vs hourly wage data later on it looks like. Pay attention to the wage symbol. 
#SAVE the data
rda_file_path <- paste0(target_folder, "ohio_region_lmi_data.rda") #rda's always better (I think?)
save(ohio_region_lmi_data, file = rda_file_path)
```


```{r IPEDS Institution Information}
#FOR FUTURE, ADD THE BELOW INSTRUCTIONS AND THE NEXT CHUNK'S INSTRUCTIONS TO A READ_ME FILE
# IPEDS Directory data -----
# https://nces.ed.gov/ipeds/use-the-data
# Survey Data > Custom Data Files
# Use provisional release data, continue
# Step 1 - Select Instituitions:
# Select "By Variables", "Browse/Search Variables"
#  Institutional Characteristics, Directory Information, select most recent year and "State abbreviation", continue, continue
#   Under "Variable Title (Table Name)" click the link "State abbreviation - (17)" and check the box for Ohio, save, submit
# Click Continue to Step 2 - Select Variables
# + Institutional Characteristics
# + Directory information, response status and frequently used variables
# + Directory information and response status:
# NOT NEEDED: Institution (entity) name -- they give you this by default, and if you request it, you get it twice!
# Institution name alias
# Street address or post office box
# City location of institution
# State abbreviation
# ZIP code
# General information telephone number
# Institution's internet website address
# Employer Identification Number
# Fips County code
# County name
# Longitude location of institution
# Latitude location of institution
# UNITID for merged schools
# Year institution was deleted from IPEDS
# Date institution closed
# Institution is active in current year
# + Institution Classifications:
# Sector of institution
# Level of institution
# Control of institution
# Highest level of offering


# Hit Continue to move on to a page listing the requested data.
# Select "STATA", which actually will produce a CSV but uses codes instead of value labels, which is good.
# .do files are also provided for each, should there be any question about value labels.
# Get JOR codes to attach to the IPEDS directory data
load('data/cross-walks/jobsohioregions.rda')

ipeds_directory <- read_csv('data/ipeds-institution-detail/STATA_RV_7162021-493.zip') %>% 
  left_join(jobsohioregions, by = c('countycd' = 'statefips')) %>% 
  transmute(
    ipeds_code = unitid,
    institutionname = instnm,
    street_address = addr,
    city = city,
    state = stabbr,
    zip = zip,
    web_address = webaddr,
    regionId = jobsohioregion,
    lat = latitude,
    lng = longitud
  )

save(ipeds_directory, file = 'data/ipeds-institution-detail/ipeds_directory.rda')
```


```{r IPEDS Completers}
# IPEDS Data -----
# https://nces.ed.gov/ipeds/use-the-data
# Survey Data > Custom Data Files
# Step 1 - Select Instituitions:
# Select "By Variables", "Browse/Search Variables"
#  Institutional Characteristics, Directory Information, select most recent year and "State abbreviation", continue, continue
#   Under "Variable Title (Table Name)" click the link "State abbreviation - (##)" and check the box for Ohio, save, submit
# Click Continue to Step 2 - Select Variables
# For each year wanted under Available Year(s) do:
#   click the year (each year refers to school year ending June 30 of that year)
#   Completions, Awards/degrees conferred by program (CIP), check Grand total
#   (repeat)
# Hit Continue to move on to a page listing each of the requested data sets.
# For each one, select "STATA", which actually will produce CSVs but uses codes instead of value labels, which is good.
# .do files are also provided for each, should there be any question about value labels.



#Usefull to get the remappings for award level, ie. these are the education levels. 
## From the STATA .do file from IPEDS for 1997:
# label values cipcode         label_cipcode        
# label define label_awlevel         15 "Degrees/certificates total"
# label define label_awlevel         12 "Degrees total", add
# label define label_awlevel         3 "Associate''s degree", add
# label define label_awlevel         5 "Bachelor''s degree", add
# label define label_awlevel         7 "Master''s degree", add
# label define label_awlevel         9 "Doctor''s degree", add
# label define label_awlevel         10 "First-professional degree", add
# label define label_awlevel         13 "Certificates below the bacculaureate total", add
# label define label_awlevel         1 "Award of less than 1 academic year", add
# label define label_awlevel         2 "Award of at least 1 but less than 2 academic years", add
# label define label_awlevel         4 "Award of at least 2 but less than 4 academic years", add
# label define label_awlevel         14 "Certificates above the bacculaureate total", add
# label define label_awlevel         6 "Postbaccalaureate certificate", add
# label define label_awlevel         8 "Post-master''s certificate", add
# label define label_awlevel         11 "First-professional certificate", add

#Using the above category definitions from the STATA file you can download from IPEDS, let's remap to levels
#so we actually know what is goin on
ipeds_degree_remapping <- tribble(
  ~awlevel, ~degree_group_logord,
  '1',        1L,
  '2',        1L,
  '3',        2L,
  '4',        1L,
  '5',        3L,
  '6',        1L,
  '7',        4L,
  '8',        5L,    # grad certificate, has not been included in the Supply Tool
  '9',        4L,
  '10',     	4L,
  '11',     	5L,    # grad certificate, has not been included in the Supply Tool
  '12',       NA,    # subtotals
  '13',       NA,    # subtotals
  '14',       NA,    # subtotals
  '15',       NA,    # subtotals
  '17',     	4L,
  '18',     	4L,
  '19',     	4L
)
  
# Read files, keep only 6-digit CIP, address some variable name changes (crace24/ctotalt)
# Using default character because it is easier to start from there, keep CIP codes correct,

#First, use list.files to find the .zip files that download from IPEDS, better to store them as .zip, because they are large. If we downloaded as above, then we should only have Ohio. 
ipeds_completions <- list.files('data/ipeds-completions', '.*zip$', full.names = TRUE) %>%  
  map_dfr(~ read_csv(., col_types = cols(.default = col_character()))) %>%
  filter(nchar(cipcode) == 7) %>%  # 7 because of the "." in the number, e.g. "15.0101"
  mutate(grads = as.integer(ctotalt)) %>% #this is the grads count column
  left_join(ipeds_degree_remapping, by = 'awlevel') %>%
  filter(!is.na(degree_group_logord) & grads > 0) %>% # drop subtotals and zero rows
  group_by(unitid, year, cipcode, degree_group_logord) %>%  # this is for combining majornum = 1 and majornum = 2...i.e we count both majors for supply. unitid is the institution. 
  summarise(graduates = sum(grads), .groups = 'drop') %>%
  left_join(transmute(ipeds_directory, unitid = as.character(ipeds_code), regionId), by = 'unitid') %>%
  select(ipeds_code = unitid, 
         cip_code = cipcode, 
         degree_group_logord, 
         academic_year = year, 
         jobsohioregion = regionId, 
         graduates)

save(ipeds_completions, file = 'data/ipeds-completions/ipeds_completions.rda')
```

##End OF data Import, now need to Combine according to CIP-SOC Crosswalk

Final Datasets Created:

ohio_region_lmi_data: Occupation demand dataset that includes six Ohio regions and statewide data (jobsohioregion coded numerically for each region).

Main Variables:
-soc_code: Standard Occupational Classification code.
-soc_lmi_title: Occupation title based on LMI.
-employment: Employment count for 2020.
-projected_2030: Projected employment count for 2030.
-change_employment: Change in employment from 2020 to 2030.
-percent_change: Percentage change in employment.
-annual_openings_growth: Annual growth in job openings.
-median_wage: Median wage in 2021.
-jobsohioregion: Region identifier (1-6 for regions, 39 for Ohio).


ipeds_completions.rda:IPEDS completions data for institutions in Ohio, linked to LMI regions.

Main Variables:
-ipeds_code: Unique identifier for institutions.
-cip_code: Classification of Instructional Programs code for program areas.
-degree_group_logord: Ordinal representation of degree levels (e.g., 1 for -certificates, 2 for associate degrees, 3 for bachelor’s degrees).
-academic_year: Year of data collection.
-jobsohioregion: Region identifier linked to LMI regions.
-graduates: Number of graduates in a given program and year.








## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
